\documentclass[12pt,twoside]{article}
\usepackage{amsmath}
\usepackage{hyperref}
\date{\today}
\title{Notes on Transport Methods in General}
\author{Rachel Slaybaugh}

\newcommand{\vOmega}{\ensuremath{\hat{\Omega}}}
\newcommand{\ve}[1]{\ensuremath{\mathbf{#1}}}


\begin{document}
\maketitle

\section*{Quasi-Diffusion}

Miften and Larsen summary \cite{Miften1993}

Miften and Larsen re-derive the quasi-diffusion (QD) equations that has an "Eddington factor" that is invariant under a change in amplitude of $\psi$ and independent of $\psi$ when it is a linear function of $\mu$. The derivation they go through doesn't appear to hold is scattering is anything other than isotropic.

The ultimate contributions are
\begin{enumerate}
\item a derivation in spherical coordinates that guarantees positivity of the analytic solution
\item new diffusion boundary conditions in planar and spherical that are more accurate and efficient
\item discretization of the new analytic forms of the QD using short characteristics, which helps ensure positivity.
\end{enumerate} 

An interesting thing they do is apply a shift in the power iteration method for the eigenvalue. They subtract a perturbation of the smaller of the absorption cross section OR the fission term from each side. 

They also do a modified QD where they have an adjustable tuning parameter that's based on quadrature, $c$ and $\sigma_T h$ - turns out the Russians did this earlier (Anistratov comment). 

In general this is all just an acceleration method, so some of the math steps are pretty kludgey. 


\vspace*{2em}
The original work is by Gol'din \cite{Goldin1964}; it's old and I think the Miften and Larsen paper is much clearer. 

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{$\alpha$ eigenvalues}

What follows is from Hill~\cite{Hill1983}, \cite{Dahl2006}, and \cite{Shen2015} (piecing together the papers, I think I've got this right...).

The time absorption, or $\alpha$, eigenvalue is an explicit eigenvalue problem. 
We begin with the time dependent TE, assume a time separability for the flux, substitute that in to convert to the time independent $\alpha$-eval form of the TE, and simultaneously add a fictitious intermediate eigenvalue (note that the flux is a vector over energy; we are using the multigroup assumption):
\begin{align*}
\ve{V}^{-1} \frac{\partial \psi}{\partial t} &+ (\ve{L} + \ve{\Sigma_t}) \psi(\vec{r}, \vOmega, t) = (\ve{S} + \ve{F})\psi(\vec{r}, \vOmega, t)\\
\psi(\vec{r}, \vOmega, t) &= \exp^{\alpha t} \psi(\vec{r}, \vOmega)\\
%
\ve{L} \psi(\vec{r}, \vOmega) &+ (\ve{\Sigma_t} + \alpha \ve{V}^{-1}) \psi(\vec{r}, \vOmega) = (\ve{S} + \frac{1}{\lambda}\ve{F}) \psi(\vec{r}, \vOmega)
\end{align*}
%
here
\begin{align*}
\ve{L} &=\text{ leakage operator}\\
\ve{\Sigma_t} &=\text{ diagonal matrix of group total cross sections}\\
\ve{V} &=\text{ diagonal matrix of neutron group speeds}\\
\ve{S} &=\text{ scattering operator}\\
\ve{F} &=\text{ fission operator}\\
\lambda &=\text{ is an intermediate eigenvalue}
\end{align*}
We treat this like a $k$-eigenvalue problem, recognizing that $\alpha$ is unknown (if $\alpha=0$, then $\lambda = k_{eff}$). 

We search for a solution set, $\alpha$ and $\vec{\psi}$, that gives $\lambda = 1$.
Inner iterations are performed for a single group on the within-group scattering source.
Outer iterations are performed on the fission source, giving the intermediate eigenvalue.
We treat angle with a discrete ordinates approximation and space with something like a finite difference method. 

The $\alpha$ eigenvalue problem is solved as a sequence of $\lambda$ eigenvalue calculations until an $\alpha$ for which $\lambda = 1$ is found. 
We perform power iteration (or equivalent) to find each $\lambda$ value and then find $\alpha$. 
(Note, you cannot use PI to find $\alpha$ directly because the fundamental mode does not correspond to the largest eigenvalue).
The thing that varies between $\alpha$ eigenvalue methods is how you then choose $\alpha^{k+1}$ based on $\alpha^k$, $\lambda^k$, and $\psi^k$ using $\alpha$ iteration index $k$.

To find each $\lambda^k$, we do power iteration. 
In general, this requires a guess for the first $\alpha$, usually $\alpha=0$, and sometimes the second (depending on method). 
We also need a starting guess for $\lambda$ and $\psi$. Thus, we get an updated $\psi$ and $\lambda$ as:
\begin{align*}
\ve{L} \psi^k &+ (\ve{\Sigma_t} - \alpha^{k-1} \ve{V}^{-1}) \psi^k -\ve{S}\psi^k = \frac{1}{\lambda^{k-1}} \ve{F} \psi^{k-1}\\
\lambda^k &= \lambda^{k-1}\frac{\ve{F} \psi^k}{\ve{F} \psi^{k-1}}\:,
\end{align*}
which is done until 
\[
|\frac{\lambda^k}{\lambda^{k-1}} - 1 | \leq \epsilon^k\:.
\]

Now, we use the updated $\lambda$ to find $\alpha^k$ in hopes of getting a $\lambda^{k+1}$ that is closer to 1. 
This $\alpha$ iteration overlying the outer and inner iterations is
\[
\ve{L} \psi^k(\vec{r}, E, \vOmega) + (\ve{\Sigma_t} + \alpha^k \ve{V}^{-1}) \psi^k(\vec{r}, E, \vOmega) = (\ve{S} + \frac{1}{\lambda^k}\ve{F}) \psi^k(\vec{r}, E, \vOmega)
\]
The thing that varies between $\alpha$ eigenvalue methods is how you then choose $\alpha^{k+1}$ based on $\alpha^k$, $\lambda^k$, and $\psi^k$ or sometimes you use further past iterates. 
Lots of  papers on this.

Dahl talks about using DSA~\cite{Dahl2006}; Fichtl talks about a non-linear method~\cite{Fichtl}; Shen has a good lit review near the opening~\cite{Shen2015}.


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Antithetic Variance}
From Carter and Cashwell's ``Particle-transport simulation with the Monte Carlo method" \href{http://www.osti.gov/scitech/biblio/4167844}{http://www.osti.gov/scitech/biblio/4167844}

Random variables $x$ and $y$ with means $\bar{x}$, $\bar{y}$ have covariance $cov(x,y) = E[(x - \bar{x})(y - \bar{y})]$ where $E[z]$ is the expected value of $z$. 
If $x$ and $y$ are independent, $cov(x,y)=0$.
The correlation coefficient of $x$ and $y$ is defined as
\[
\rho(x,y) = \frac{cov(x,y)}{[var(x)var(y)]^{1/2}}\:,
\]
which has a value between -1 and 1 (Cauchy-Schwarz inequality).
If $\rho=0$, the variables are said to be uncorrelated, if $>0$ positively correlated, if $<0$ negatively correlated.

We have an estimator $t$ with expectation $\theta$. 
We seek estimator $t*$ with the same expectation but a negative correlation. 
Then $(t+t*)/2$ will be an unbiased estimator of $\theta$ with variance
\[
var \bigl[\frac{(t + t*)}{2} \bigr] = \frac{var(t)}{4} + \frac{var(t*)}{4} + \frac{cov(t,t*)}{2}\:.
\]
If $cov(t,t*) < 0$, then maybe 
\[var\bigl[\frac{(t + t*)}{2}\bigr] < var(t)\:.\]

They then go through some examples illustrating this concept. 

Prove that given two monotone nondecreasing (nonincreasing) functions $g(x)$ and $q(x)$ with 
\[
\int_0^1 q(x) dx = 0
\]
Then
\[
\int_0^1 g(x)q(x) dx \geq 0
\]
They also have defined a strictly monotone estimator $g(x)$ and a corresponding
\[
f(x) = \frac{g(x) + g(1-x)}{2}\:,
\]
calling $V = var(g)$ and $v_1 = var(f)$. 
They use the definition of $f(x)$ in combination with the proof to demonstrate that 
\[
V_1 \leq \frac{V}{2}\:.
\]
They further mention the case for when $g(x)$ is not necessarily monotone. 

``As a simple example of the use of antithetic variates  in a practical situation, Hammersley and Handcomb discuss the problem of calculating the  multiplication rate in a sphere of  fissile material, where it is assumed that fission neutrons are emitted isotropically. In estimating the eigenvalue, they chose that neutrons in pairs from fission  be emitted in opposite directions and made estimates of the expected number of  fissions occurring in the system in each of the two directions. These estimates are negatively correlated since, if one is too large, then the other is too low. A better estimate is then expected to be the arithmetic mean  of the two."  

``Another example mentioned by Spanier and Gelbard involves the use of antithetic variates in estimating resonance integrals. Neutrons born at higher  energies will usually contribute more heavily to the resonance integrals  simply  because they pass through more energies at which  resonances occur. The source of resonance neutrons is sampled in such a manner that a particle born at high energy is paired with a particle born at a correspondingly low energy. These two  contributions are negatively correlated, and using their mean as the value of the contribution can lead to much more efficient sampling of the resonance integrals."


%------------- Bibliography --------------------
\bibliographystyle{siam}
\bibliography{te_general}

\end{document}